Concurrent Programming with Harmony

### 2020-08-19 book/paper.tex
Latest

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queue}
uses the \texttt{alloc} module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using \texttt{malloc}() and \texttt{free}().
\texttt{malloc}() returns a new memory location, which should be released
with \texttt{free}() when it is no longer in use.
The queue maintains a pointer to the first and last elements in the list,
which are \textbf{None} if the queue is empty.

\texttt{Qnew()} returns a new queue object consisting of a \textbf{None} head
and tail pointer and a lock.
\texttt{Qenqueue(q, v)} and \texttt{Qdequeue(q)} both take a pointer $q$ to the
queue object because both may modify the queue.
Before they access the value of the head or tail of the queue they first obtain
the lock.
When they are done, they release the lock.

The implementation in \autoref{fig:queueMS}~\cite{MS98} is similar but uses
separate locks for the head and tail, allowing an enqueue and dequeue operation
to proceed concurrently.  To avoid contention between the head and the tail, the
queue uses a dummy item on the linked list.

@@

@@

\section*{Exercises}
\begin{problems}
\item Draw additional states and steps in \autoref{fig:rdwr}
for three processes.
\item Optimize your solution to \autoref{ex:xy} to use reader/writer locks.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} processes to acquire it at the same time.  A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
\item Write a test program for your bound lock
that checks that no more than \texttt{M} processes can acquire the
bound lock.
\end{problems}

@@

\begin{figure}
\begin{code}
\harmonysource{semaphore}
\end{code}
\caption{The \texttt{Semaphore} interface in the \texttt{synch} module.}
\label{fig:semaphore}
\end{figure}

@@

\begin{figure}
\begin{code}
\harmonysource{mesa}
\end{code}
\caption{Implementation of condition variables in the \texttt{synch} module.}
\label{fig:cv}
\end{figure}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{malloc()} & allocate and return a pointer to a memory location \\
\hline
\texttt{free($p$)} & free an allocated memory location $p$ \\
\hline
\end{tabular}
\vspace{1em}

>>>>

We will first implement the queue as a linked list.
The implementation in \autoref{fig:queue}
uses the \texttt{alloc} module for dynamic allocation
\index{dynamic allocation}%
of nodes in the list using \texttt{malloc}() and \texttt{free}().
\texttt{malloc}($v$) returns a new memory location initialized to $v$,
which should be released with \texttt{free}() when it is no longer in use.
The queue maintains a \texttt{head} pointer to the first element in the list
and a \texttt{tail} pointer to the last element in the list.
The \texttt{head} pointer is \textbf{None} if and only if the queue is empty.

\texttt{Qnew()} returns a new queue object consisting of a \textbf{None} head
and tail pointer and a lock.
$\mathtt{Qenqueue}(q, v)$ and $\mathtt{Qdequeue}(q)$ both take a pointer $q$ to the
queue object because both may modify the queue.
Before they access the value of the head or tail of the queue they first obtain
the lock.
When they are done, they release the lock.

The implementation in \autoref{fig:queueMS}~\cite{MS98} is similar but uses
separate locks for the head and tail, allowing an enqueue and a dequeue operation
to proceed concurrently.  To avoid contention between the head and the tail that
requires both locks, the queue uses a dummy node at the head of the linked list.
Except of the initial state of the queue, the dummy node is the last node that
was dequeued.  Note that neither the \texttt{head} nor \texttt{tail} pointer are
ever \textbf{None}.

@@

\section*{Exercises}
\begin{problems}
\item \label{ex:qcontains} Add a method $\mathtt{Qcontains}(q, v)$ to both \autoref{fig:queue} and \autoref{fig:queueMS}
that checks to see if $v$ is in queue $q$.
\item Add a method $\mathtt{Qremove}(q, v)$ to both \autoref{fig:queue} and \autoref{fig:queueMS}
that removes $v$, if any, from queue $q$.
\item The test program in \autoref{fig:queuetest} is not very thorough.  Design
and implement a better one.
\item Create a thread-safe sorted binary tree.  Implement methods
$\mathtt{BTnew}()$ to create a new binary tree, $\mathtt{BTinsert}(t, v)$ that inserts $v$ into
tree $t$, and $\mathtt{BTcontains}(t, v)$ that checks if $v$ is in tree $t$.  Use a single
lock per binary tree.
\item Bonus question: create a binary tree that uses, instead of a single lock per tree,
a lock for each node in the tree.
\end{problems}

@@

\section*{Exercises}
\begin{problems}
\item Draw additional states and steps in \autoref{fig:rdwr}
for three processes.
\item Optimize your solutions to \autoref{ex:qcontains} to use reader/writer locks.
\item \label{ex:boundlock} Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} processes to acquire it at the same time.  A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
\item Write a test program for your bound lock
that checks that no more than \texttt{M} processes can acquire the
bound lock.
\end{problems}

@@

\begin{figure}
\begin{code}
\harmonysource{semaphore}
\end{code}
\caption{\harmonylink{modules/synch.hny} The \texttt{Semaphore} interface in the \texttt{synch} module.}
\label{fig:semaphore}
\end{figure}

@@

\begin{figure}
\begin{code}
\harmonysource{mesa}
\end{code}
\caption{\harmonylink{modules/synch.hny}
Implementation of condition variables in the \texttt{synch} module.}
\label{fig:cv}
\end{figure}

@@

\vspace{1em}
\begin{tabular}{|l|l|}
\hline
\texttt{malloc($v$)} & return a pointer to a memory location initialized to $v$ \\
\hline
\texttt{free($p$)} & free an allocated memory location $p$ \\
\hline
\end{tabular}
\vspace{1em}
